---
title: "MSstatsLiP_TimeSeries"
author: "Devon Kohler (<kohler.d@northeastern.edu>)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MSstatsTMTPTM Example: An example workflow and analysis of the MSstatsTMTPTM package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8, 
  fig.height=8
)
```

## MSstatsLiP Time Series Vignette

This Vignette provides an example workflow for how to model a time series 
experiment using MSstatsLiP.

```{r message=FALSE, warning=FALSE}
library(MSstatsLiP)
library(tidyverse)
```

## Workflow

### 1. Preprocessing

### 1.1 Raw Data Format

The first step is to load in the raw dataset for both the PTM and Protein 
datasets. This data can then be converted into MSstatsLiP format with one of the
built in converters, or manually converted.

The experimental design is critical when modeling time series experiments. The 
package will automatically detect whether to model the time series by how the 
design is entered into the raw files.

The table below shows the values of the columns Condition, BioReplicate and Run 
for a time series experiment. Note the same subject is repeated for different 
time points (Conditions).

| Condition | BioReplicate | Run |
|:---------:|:------------:|:---:|
| Time1     | Subject1     | 1   |
| Time2     | Subject1     | 2   |
| Time1     | Subject2     | 3   |
| Time2     | Subject2     | 4   |
| Time1     | Subject3     | 5   |
| Time2     | Subject3     | 6   |

### 1.2 Data Example

``` {r}
## Include time series data in final package?
#head(LiPRawData)
#head(TrPRawData)

raw_lip <- read.delim("F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/210124_FE-413_activity_markers_LiP_Report.csv", sep = ',')
raw_prot <- read.delim("F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/210124_FE-413_activity_markers_Trp_Report.csv", sep = ',')

raw_lip_sel <- raw_lip  %>%
  filter(str_detect(R.Condition,"IPTG"))
raw_prot_sel <- raw_prot  %>%
  filter(str_detect(R.Condition,"IPTG"))
raw_prot_sel$R.Condition = substr(raw_prot_sel$R.Condition,1,nchar(raw_prot_sel$R.Condition)-4) 
raw_lip_sel$R.Condition = substr(raw_lip_sel$R.Condition,1,nchar(raw_lip_sel$R.Condition)-3) 

## Sample proteins to increase speed
sample <- raw_lip_sel %>% filter(!grepl(";", PG.ProteinAccessions)) %>% distinct(PG.ProteinAccessions) %>% sample_n(250)
raw_lip_sel <- raw_lip_sel %>% filter(PG.ProteinAccessions %in% sample[[1]])
raw_prot_sel <- raw_prot_sel %>% filter(PG.ProteinAccessions %in% sample[[1]])


MSstatsLiP_data <- SpectronauttoMSstatsLiPFormat(raw_lip_sel, 
    "F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/200331_ecoli.fasta",
                                                 raw_prot_sel)

head(MSstatsLiP_data[["LiP"]])
head(MSstatsLiP_data[["TrP"]])

```

To confirm the data is in the correct format we can take a look at the 
Condtion, Bioreplicate, and Run columns.

```{r}
MSstatsLiP_data[["TrP"]] %>% 
  distinct(BioReplicate,Condition, Run) %>% 
  arrange(BioReplicate, Condition, Run)
```

### 2. Summarize

``` {r}

MSstatsLiP_Summarized <- dataSummarizationLiP(MSstatsLiP_data,
                                              MBimpute = FALSE)
names(MSstatsLiP_Summarized[["LiP"]])
head(MSstatsLiP_Summarized[["LiP"]]$RunlevelData)
head(MSstatsLiP_Summarized[["TrP"]]$RunlevelData)

```

### 3. Modeling

The modeling function groupComparisonLiP takes as input the output of the 
summarization function dataSummarizationLiP.

Because this is a time series it makes more sense to define out comparisons of 
interest, rather than perform a full pairwise comparison.

```{r}

contrast_matrix <- matrix(c(-1,1,0,0,0,0,0,0,
                         -1,0,1,0,0,0,0,0,
                         -1,0,0,1,0,0,0,0,
                         -1,0,0,0,1,0,0,0,
                         -1,0,0,0,0,1,0,0,
                         -1,0,0,0,0,0,1,0,
                         -1,0,0,0,0,0,0,1),
                         nrow=7, ncol=8, byrow=TRUE)

row.names(contrast_matrix) <- c("T1-T2", "T1-T3", "T1-T4", 
                                "T1-T5", "T1-T6", "T1-T7", "T1-T8")

MSstatsLiP_model <- groupComparisonLiP(MSstatsLiP_Summarized, contrast_matrix,
  fasta = "F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/200331_ecoli.fasta")

head(MSstatsLiP_model[["LiP.Model"]])
head(MSstatsLiP_model[["TrP.Model"]])
head(MSstatsLiP_model[["Adjusted.LiP.Model"]])

```

#### 3.1 Volcano Plot

``` {r}
groupComparisonPlotsLiP(MSstatsLiP_model, 
                        type = "VolcanoPlot", 
                        ProteinName = FALSE, 
                        address = FALSE)
```
