---
title: "MSstatsLiP_Workflow"
author: "Devon Kohler"
date: "March 4, 2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MSstatsLiP Workflow Vignette

This Vignette provides an example workflow for how to use the package 
MSstatsLiP.

## Installation

To install this package, start R (version "4.0") and enter:

``` {r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MSstatsTMTPTM")
```

```{r}
library(MSstatsLiP)

library(MSstatsPTM)
library(tidyverse)
library(MSstats)
library(data.table)
```
## 1. Workflow

### 1.1 Raw Data Format

The first step is to load in the raw dataset for both the PTM and Protein 
datasets. This data can then be converted into MSstatsLiP format with one of the
built in converters, or manually converted. In this case we use the converter 
for Spectronaut

### 1.2 Converter

``` {r}
# read in raw data files
load(file = "../data/Example_Lip.rda")
load(file = "../data/Example_Trp.rda")

msstats_data <- SpectronauttoMSstatsLiPFormat(raw_lip, raw_prot,
        "F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/200331_ecoli.fasta")

head(msstats_data[["LiP"]])
head(msstats_data[["TrP"]])

## Make conditions match
msstats_data[["TrP"]]$Condition = substr(msstats_data[["TrP"]]$Condition,1,nchar(msstats_data[["TrP"]]$Condition)-4) 
msstats_data[["LiP"]]$Condition = substr(msstats_data[["LiP"]]$Condition,1,nchar(msstats_data[["LiP"]]$Condition)-3) 
```

#### 1.2.2 Barcode plot

```{r}
BarcodePlotLiP(msstats_data, "F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/200331_ecoli.fasta", 
               which.prot = c("Q9JMR6", "A5A614"))

```


The converter function returns a list of two dataframes one each for LiP and TrP
data. This list is the input into the next summarization step.

### 1.3 Summarization

#### 1.3.1 Function

``` {r}

msstats_summarized <- dataSummarizationLiP(msstats_data)
names(msstats_summarized[["LiP"]])
head(msstats_summarized[["LiP"]]$ProcessedData)
head(msstats_summarized[["LiP"]]$RunlevelData)
head(msstats_summarized[["TrP"]]$ProcessedData)
head(msstats_summarized[["TrP"]]$RunlevelData)

save(msstats_summarized, file = '../data/summarized_temp.rda')
load(file = '../data/summarized_temp.rda')
```

Again the summarization function returns a list of two dataframes one each for 
LiP and TrP. Each LiP and TrP is also a list with additional summary 
information. This summarized data can be used as input into some of the
plotting functions included in the package.

#### 1.3.2 QCPlots

``` {r}
## Profile Plots
dataProcessPlotsLiP(msstats_summarized,
                  type = 'QCPLOT',
                  which.Protein = "allonly",
                  address = FALSE,
                  lip.title = "All LiP Peptides")

dataProcessPlotsLiP(msstats_summarized,
                    type = 'ProfilePlot',
                    which.Protein = 1,
                    ylimUp=FALSE,
                    ylimDown=10,
                    address = FALSE)
  

```

#### 1.3.3 PCA Plot

``` {r}

PCAPlotLiP(msstats_summarized,
                       center.pca = TRUE,
                       scale.pca = TRUE,
                       n.components = 10,
                       bar.plot = TRUE,
                       protein.pca = FALSE,
                       comparison.pca = TRUE,
                       which.pep = "all",
                       which.comparison = c("IPTG_1_1", "IPTG_1_2"
                                            , "IPTG_2_2", "IPTG_2_2"),
                       width=10,
                       height=10,
                       address=FALSE)

```

### 1.4 Modeling

#### 1.4.1 Function

```{r}

msstats_model <- groupComparisonLiP(msstats_summarized)

head(msstats_model[["LiP.Model"]])
head(msstats_model[["TrP.model"]])
head(msstats_model[["Adjusted.LiP.model"]])

```

#### 1.4.2 Plots

``` {r}
load(file = "F:/Northeastern/Research/MSstats/MSstatsLiP/Ecoli_Val/Ecoli_model_IPTG.rda")

groupComparisonPlotsLiP(msstats_model, 
                        type = "VolcanoPlot", 
                        ProteinName = FALSE, 
                        which.Comparison = "IPTG_1-IPTG_4",
                        address = FALSE)

groupComparisonPlotsLiP(LipExpt.model, type = "HEATMAP",
                        which.Comparison = c("IPTG_1-IPTG_2","IPTG_1-IPTG_3","IPTG_1-IPTG_4", "IPTG_1-IPTG_5"),
                        numProtein=50,
                        address = "",
                        which.Protein=1:50)

```
